Efficient pedigree recording for fast population genetics simulation

### 2018-01-09 forwards_paper.tex
Light editing pass + added some references.

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This allows us to dramatically reduce the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history in the \emph{tree sequence} format introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one or two orders of magnitude.
    In addition to speed, we show that the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored on $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation
    rather than only a subset.
    (3) A simulation can easily be initialized with prior history produced by a more efficient coalescent simulation of deep history.
    To make these possible, 
    we introduce the \msprime{} Tables API, which allows for the efficient interchange of genealogical data between modular program components.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \citep{Neuhauser1997-nn, hudson1990gene},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

@@

We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

@@

The simulations above recorded information about each new individual
in collection of tables that together encode a \emph{tree sequence}.
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
A closely related structure is known as the \emph{ancestral recombination graph} --
the {ARG} \citep{griffiths1991two,griffiths1997ancestral},
where nodes are common ancestor or recombination events --
which has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, the properties of the ARG as a computational structure have not
been studied and, despite several efforts to standardise a common
format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper.
ARGs are rarely used
in practise. In contrast, the algorithmic properties of tree sequence
algorithms have been explored in detail~\citep{kelleher2016efficient},
contributing substantially to the efficiency of the \msprime\ coalescent simulator.

@@

The \emph{edges} define how nodes relate to each other over specific genomic intervals.
Each edge records:
% is a tuple $(\ell, r, p, c)$, where
the endpoints $[\ell, r)$ of the half-open genomic interval defining the
spatial extent of the edge;
and the identities $p$ and $c$ of the parent and child nodes
of a single branch that occurs in all trees covering this interval.
The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
are shown in the bottom left panel.
For example, the branch joining nodes 1 to 3 appears in both trees,
and so is recorded as a single edge extending over the whole chromosome.
It is this method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding very compact and algorithmically efficient.

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 with ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
Each mutation occurs above a specific node at a given site,
and results in a specific derived state.
Thus, all samples below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red hexagons.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

@@

\algstep{W4.}{Record edges}{
Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, r, P_b, u}$.
}

\algstep{W5.}{Individual loop}{Set $j \leftarrow j + 1$. If $j < N$ go to \algref{W2}.
Otherwise, if $t\bmod s \neq 0$ go to \algref{W7}.
}

@@

Figure~\ref{fig:wf-trees} shows
an example of a marginal genealogy produced by a forwards-time Wright--Fisher
process like Algorithm~\algref{W}.
On the left is the tree showing all the edges output by the simulation,
while on the right
is the minimal tree representing the ancestry of the current set of samples.
Clearly there is a great deal of redundancy in the topological
information output by the simulation.
This redundancy comes from two sources.
First, there are a large number of nodes in tree that have only one child.
In Algorithm~\algref{W} we do not attempt to identify coalescence events,
but simply record all parent-child
relationships in the history of the population.
As such, many of these edges
will record the simple passing of genealogical information from parent to child
and only some small subset will correspond to coalescences within the marginal
trees. The second source of redundancy in the output of Algorithm~\algref{W}
is due to the fact that lineages die out: a large number of
individuals in the simulation leave no descendants in the present day population.
Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
ancestors in the current population, and so the entire path tracing back to
the root is redundant.

@@

Any simulation scheme that records data into tables,
as does Algorithm~\algref{W},
has its genealogical history available at any time as a tree sequence.
This has two additional advantages:
First, simplification can be run periodically through the simulation,
if we take the set of samples to be the entire currently alive population.
This is important in practice as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime,
providing an easy, efficient way to specify prior history.
A natural question is now: how often should simplification occur?  We did some limited testing with the \cpp{}
implementation of our simulations, and found that different simplification intervals affect run times by approximately
25\%, with the lowest run time occuring when simplifying every $10^3$ generations.  Thus, there is a memory-versus-speed
tradeoff as simplifying more often would keep fewer extinct nodes and edges in memory.  The run-time costs associated
with simplifying have to do with preparing the data to be sent to the simplification algorithm, rather than with the
algorithm itself, which we show in the next section has linear time complexity.

@@

The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
near-linear in $rho$ (Figure \ref{fig:runtimes_selection}).  In this section, we explore the run-time complexity using a
simplified Wright-Fisher model with recombination.

@@

The Tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program.
% we are capitalizing Tables API but not the more generic "tables".
% \jda{should capitalize Tables, or not, as above}
Following current best practises
\plr{citations: Apache arrow, etc},
data is stored in a columnar format,
which allows for very efficient interchange of large amounts of numerical data.
In principle, this enables zero-copy semantics,
where a data consumer can read the information directly
from the memory of a producer without incurring the overhead of a copy.
\jda{TODO citation missing}
Our implementation uses the numpy C API \citep{numpy_C_api} to efficiently copy
data from Python into the low-level C library used to manipulate tree sequences.
Since tables are encoded as contiguous arrays of numeric data,
this achieves data transfer rates that would be impossible under
any text-based approach while retaining excellent portability.

This also provides basic input and output operations via numpy's
array interface, which provides a great deal of flexibility and efficiency,
and makes it straightforward to transfer data from sources
such as HDF5 \citep{hdf5}, Dask, or Zarr.
\jda{TODO citation missing}
(For small scale data and debugging purposes, a simple text format is also supported.)
% PLR: this was mentioned above, and belongs in the documentation, I think.
% The API also provides functions to sort tables
% to ensure that the records are in the form required to input
% into an \msprime{} tree sequence object. (Simplification may also be required.)

>>>>

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This dramatically reduces the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history in the \emph{tree sequence} format introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one or two orders of magnitude.
    In addition to speed, we show that the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored on $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation
    rather than only a subset.
    (3) A simulation can easily be initialized with prior history produced by a more efficient coalescent simulation of deep history.
    To make these possible,
    we introduce the \msprime{} Tables API, which allows for the efficient interchange of genealogical data between modular program components.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical
space~\citep{barton2010new,kelleher2014coalescent}
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples approaching the size of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

@@

We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

@@

The simulations above recorded information about each new individual
in a collection of tables that together define a \emph{tree sequence}.
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
A closely related structure is known as the \emph{ancestral recombination graph} --
the {ARG} \citep{griffiths1991two,griffiths1997ancestral},
where nodes are common ancestor or recombination events --
which has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, the properties of the ARG as a computational structure have not
been studied and, despite several efforts to standardise a common
format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper.
ARGs are rarely used
in practise. In contrast, the algorithmic properties of tree sequence
algorithms have been explored in detail~\citep{kelleher2016efficient},
contributing substantially to the efficiency of the \msprime\ coalescent simulator.

@@

The \emph{edges} define how nodes relate to each other over specific genomic intervals.
Each edge records:
% is a tuple $(\ell, r, p, c)$, where
the endpoints $[\ell, r)$ of the half-open genomic interval defining the
spatial extent of the edge;
and the identities $p$ and $c$ of the parent and child nodes
of a single branch that occurs in all trees in this interval.
The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
are shown in the bottom left panel.
For example, the branch joining nodes 1 to 3 appears in both trees,
and so is recorded as a single edge extending over the whole chromosome.
It is this method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding compact and algorithmically efficient.

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
Each mutation occurs above a specific node at a given site,
and results in a specific derived state.
Thus, all samples below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red stars.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

@@

\algstep{W4.}{Record edges.}{
Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, r, P_b, u}$.
}

\algstep{W5.}{Individual loop.}{Set $j \leftarrow j + 1$. If $j < N$ go to \algref{W2}.
Otherwise, if $t\bmod s \neq 0$ go to \algref{W7}.
}

@@

Figure~\ref{fig:wf-trees} shows
an example of a marginal genealogy produced by a forwards-time Wright--Fisher
process like Algorithm~\algref{W}.
On the left is the tree showing all the edges output by the simulation,
while on the right
is the minimal tree representing the ancestry of the current set of samples.
Clearly there is a great deal of redundancy in the topological
information output by the simulation.
This redundancy comes from two sources.
First, there are a large number of nodes in the tree that have only one child.
In Algorithm~\algref{W} we do not attempt to identify coalescence events,
but simply record all parent-child
relationships in the history of the population.
As such, many of these edges
will record the simple passing of genealogical information from parent to child
and only some small subset will correspond to coalescences within the marginal
trees. The second source of redundancy in the (unsimplified) output of Algorithm~\algref{W}
is due to the fact that lineages die out: a large number of
individuals in the simulation leave no descendants in the present day population.
Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
ancestors in the current population, and so the entire path tracing back to
the root is redundant.

@@

Any simulation scheme that records data into tables,
as Algorithm~\algref{W} does,
has its genealogical history available at any time as a tree sequence.
This has two additional advantages:
First, simplification can be run periodically through the simulation,
if we take the set of samples to be the entire currently alive population.
This is important in practice as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime,
providing an easy, efficient way to specify prior history.
A natural question is now: how often should simplification occur?  We did some limited testing with the \cpp{}
implementation of our simulations, and found that different simplification intervals affect run times by approximately
25\%, with the lowest run time occuring when simplifying every $10^3$ generations.  Thus, there is a memory-versus-speed
tradeoff as simplifying more often would keep fewer extinct nodes and edges in memory.  The run-time costs associated
with simplifying have to do with preparing the data to be sent to the simplification algorithm, rather than with the
algorithm itself, which we show in the next section has linear time complexity.

@@

The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  In this section, we explore the run-time complexity using a
simplified Wright-Fisher model with recombination.

@@

The Tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program.
Following current best practises
\plr{citations: Apache arrow, etc},
data is stored in a columnar format,
which allows for very efficient interchange of large amounts of numerical data.
In principle, this enables zero-copy semantics,
where a data consumer can read the information directly
from the memory of a producer without incurring the overhead of a copy.
\jda{TODO citation missing}
Our implementation uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
data from Python into the low-level C library used to manipulate tree sequences.
Since tables are encoded as contiguous arrays of numeric data,
this achieves data transfer rates that would be impossible under
any text-based approach while retaining excellent portability.

This also provides basic input and output operations via NumPy's
array interface, which provides a great deal of flexibility and efficiency,
and makes it straightforward to transfer data from sources
such as HDF5 \citep{hdf5}or Dask~\citep{dask}.
For small scale data and debugging purposes, a simple text format is also supported.

% PLR: this was mentioned above, and belongs in the documentation, I think.
% The API also provides functions to sort tables
% to ensure that the records are in the form required to input
% into an \msprime{} tree sequence object. (Simplification may also be required.)
